Synthesizing Quotient Lenses

### 2018-02-07 papers/QuotientLenses/main.tex
Made a pass through paper

\section{Quotient Regular Expressions and Lenses by Example}
\label{sec:example}
In this section, we explain the difficulties of programming with quotient lenses
and show, by way of example, how our novel language of {\em Quotient Regular
Expressions}, or QREs, and QRE lenses address these difficulties.

@@

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge.  Three of the inessential details
we wish to ignore---whitespace quotienting
inside fields, whitespace quotienting in between fields, and permutation
quotenting of fields---nest within each other in a way that makes them difficult
to express with the existing Boomerang combinators.

Furthermore, programming best practices require programmers provide type
annotations to their lenses to make explicit the formats the quotient lenses map
between. This practice prevents nefarious bugs that can result from the lenses
mapping between incorrect formats, but it is also verbose and time consuming.

To address these issues, we propose {\em Quotient Regular Expressions}, or QREs,
which enable the programmer to define a regular expression $R$ and its canonical
elements in the same expression. For example, with QREs we can specify \bibtex{}
records and their canonical elements, as shown in Figure~\ref{fig:example-qre}.

@@

In Figure~\ref{fig:example-qre}, the QRE \verb+project [ \n\t\r]+$^{+} \to$
\verb+" "+ matches any non-zero number of whitespace characters, with a single
space as the canonical representative. We call this QRE a \textit{projection}
because an entire regular language is projected to a single representative
string during canonization.

@@

\noindent This QRE is a \textit{squash} QRE. A \textit{squash} QRE ``glues''
two data formats using a user-defined function. For example,
the QRE in the code above specifies that the \cd{comma_name} representation of
the author's name is equivalent to the \cd{space_name} representation, with
\cd{space_name} taken to be the canonical representation. The
function that identifies the two formats is given by \cd{get space_to_comma},
which transforms names in the \cd{comma_name} representation to names in the
\cd{space_name} representation. In this example, \cd{space_to_comma} is itself
an automatically \textit{synthesized} lens between the \cd{space_name} and
\cd{comma_name} QREs, and we need to use the ``get'' direction of the lens to
specify the squash transformation. (We'll say more about synthesis below.)

@@

\sam{TODO: Put diagram of composition here}

@@

As we saw with the definition of \cd{space_to_comma} in
Figure~\ref{fig:example-qre}, the synthesis procedure itself can be used to
creates lenses that are in turn used to define other QREs.  This ability to mix
QRE specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

@@

\saz{Fix the notation in the following code.}

@@

\saz{Say more about how normalize
wouldn't be used in ``ordinary'' QRE programming?}\sam{I'm not sure what you
mean here exactly\ldots} However, checking that the canonizing function $f$ is
surjective and idempotent is undecidable in general, thus we do not use the
$\normalize{R}{R'}{f}$ in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonizer{}$ function for each QRE. Again, the \textit{permutation} combinator
gives the most interesting definition,

\begin{align*}\canonizer(\perm{c_1, \ldots, c_n}{c})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonizer(c_1)(r_1)) \cdot (\canonizer(c)(s_1)) \cdot \ldots \cdot
(\canonizer(c_n)(r_n))
\end{align*}

\noindent which as we have explained is defined so that the strings that match
$c_i, c$ are placed in the canonical permutation which is the identity
permutation before $\canonizer(c_i)$ and $\canonizer(c)$ are applied
respectively.

@@

The denotation of a lens $\ell$ is $\llbracket \ell \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell
\rrbracket$, then $\ell$ maps between $s_1$ and $s_2$.

\begin{align*}
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket const(r, s) \rrbracket &= \{(r, s)\}\\
\llbracket swap(\ell, \ell') \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \cdot \ell' \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \sep \ell' \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell \rrbracket \text{ or } (s, t) \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}\\
\llbracket \ell_2 \circ \ell_1 \rrbracket &= \{(r, t) \sep \text{there exists }s
\text{ such that } (r, s) \in \llbracket \ell_1 \rrbracket \text{ and } (s, t)
\in \llbracket \ell_2 \rrbracket\}
\end{align*}

The $\mathit{const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the backward
direction. $\mathit{id}(R)$ applies the identity function to the source and target
in $\mathcal{L}(R)$ in both directions. The composition lens $\ell' \circ \ell$
applies $\ell$ followed by then $\ell'$ to the source in the forward direction,
and applies $\ell'$ followed by $\ell$ to the target in the backward direction.
The lens $\ell \cdot \ell'$ first splits the string $s$ into $s_1$ and $s_2$,
applies $\ell$ and $\ell'$ to $s_1$ and $s_2$ to get $t_1$ and $t_2$
respectively, then concatenates $t_1$ and $t_2$ and returns $t_1 \cdot t_2$ as
the final result in the forward direction. $\ell \cdot \ell'$ operates
similarly in the backward direction, but with $s, s_1$ and $s_2$ substituted
for $t, t_1$ and $t_2$. The $\mathit{swap} \; (\ell, \ell')$ lens operates
like $\ell \cdot \ell'$, except that it swaps $t_1$ and $t_2$ before
concatenating the two for a final result of $t_2 \cdot t_1$ in the forward
direction. In the backward direction, $\mathit{swap}(\ell, \ell')$ first undoes
the swap, then proceeds as expected. The $\ell \; | \; \ell'$ lens
chooses to apply $\ell$ or $\ell'$ depending on whether the source
(resp. target) data is matched by $\ell$ or $\ell'$ in the forward (resp.
backward direction). The $\ell^*$ lens splits the string $s$ into strings $s_1,
\ldots, s_n$, applies $\ell$ to each $s_i$ to get $t_i$, and then concatenates
each of the $t_i$'s for a final result of $t_1 \cdot \ldots \cdot t_n$ in the
forward direction. $\ell^*$ operates similarly in the backward direction, but
with $s, s_i$ substituted for $t, t_i$.

@@

\begin{lstlisting}
let bib_to_end_q : (lens in bibtex $\Leftrightarrow$ endnote) =
rquot (lquot (bibtex, bib_to_end), endnote)
\end{lstlisting}

@@

In general, checking that $\canonizer(c') = \canonizer(c'')$ is an undecidable
task, especially because of the $\normalize{R}{R'}{f}$ QRE. Therefore, our
actual implementation of the composition rule for QRE lenses follows Boomerang's
approach, which is to define the composition on QRE lenses if and only if
$c'$ and $c''$ are syntactially equal to $id(R)$ and $id(R')$ for some $R$and
$R'$ satisfying $\mathcal{L}(R) = \mathcal{L}(R')$.

@@

\begin{theorem}\label{normal form}
If there is a derivation $q : c \Leftrightarrow c'$, then there exists a
bijective lens $\ell : K(c) \Leftrightarrow K(c')$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c')
\end{align*}
\end{theorem}
\begin{proof}
We elide a full proof of this statement, since the proof is attained via a
straightforward induction on the derivation $q : c \Leftrightarrow c$. \saz{(The
full details can be found in the appendix?)}Since
bijective lenses are closed under strongly unambiguous regular operators, the
tricky case is composition. The composition case is more involved since we need
to show that we can get rid of the canonizers in the middle:

@@

\fi
\section{Synthesizing QRE Lenses}
\label{synth}
Our goal in this section is to demonstrate how to synthesize a quotient lens
$q: c \Leftrightarrow c'$ from the QREs $c, c'$ and a set of input-output pairs
$\{(x_1, y_1), \ldots, (x_n, y_n)\}$ where the $x_i$'s are in $W(c)$ and the
$y_i$'s are in $W(c')$. We wish $q$ to map the equivalence class of $x_i$ to the
equivalence class of $y_i$ and vice versa:
\begin{align*}
q.get(x_i) \equiv_{c'} y_i \text{, and}\\
q.put(y_i) \equiv_c x_i
\end{align*}

>>>>

\section{Quotient Regular Expressions and Lenses by Example}
\label{sec:example}
In this section, we explain the difficulties of programming with quotient lenses
in Boomerang and show, by way of example, how our novel language of {\em
Quotient Regular Expressions}, or QREs, and QRE lenses address these
difficulties.

@@

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge.  Three of the inessential details
we wish to ignore---whitespace quotienting
inside fields, whitespace quotienting in between fields, and permutation
quotenting of fields---nest within each other in a way that makes them difficult
to express with the existing Boomerang combinators. Furthermore, if the
programmer wishes to specify the type of the lens, then she can only specify
data formats on each side of the lens with a regular expression that does not
keep track of the equivalence relation defined on the language of the regular
expression. Consequently, the programmer has to write a regular expression as
well as a canonizer that expresses the equivalence relation on that regular
expression, even though a regular expression and the canonizer defined on it
typically share much of the syntactic structure. Having to write both of these
expressions makes programming with quotient lenses verbose and time consuming.

To address these issues, we propose {\em Quotient Regular Expressions}, or QREs,
which enable the programmer to define a regular expression $R$, an equivalence
relation $\sim$ on the language described by $R$ as well as a canonical set
of representatives of $\sim$, all in one syntactic term. For example, with QREs
we can specify \bibtex{} records and their canonical elements, as shown in
Figure~\ref{fig:example-qre}.

@@

In Figure~\ref{fig:example-qre}, the QRE \verb+project [ \n\t\r]+$^{+} \to$
\verb+" "+ matches any non-zero number of whitespace characters and quotients
them to a single equivalence class with a single space as the canonical
representative. We call this QRE a \textit{projection} because an entire regular
language is projected to a single representative string.

@@

\noindent This QRE is a \textit{squash} QRE. A \textit{squash} QRE ``glues''
two data formats using a user-defined function. For example,
the QRE in the code above specifies that the \cd{comma_name} representation of
the author's name is equivalent to the \cd{space_name} representation, with
\cd{space_name} taken to be the canonical representation. The
function that identifies the two formats is given by \cd{get space_to_comma},
which transforms names in the \cd{comma_name} representation to names in the
\cd{space_name} representation. In this example, \cd{space_to_comma} is itself
an automatically \textit{synthesized} lens between the \cd{space_name} and
\cd{comma_name} QREs, and we need to use the ``get'' direction of the lens to
specify the squash transformation.

@@

\sam{TODO: We need to put a diagram of composition here.}

@@

As we saw with the definition of \cd{space_to_comma} in
Figure~\ref{fig:example-qre}, the synthesis procedure itself can be used to
create lenses that are in turn used to define other QREs.  This ability to mix
QRE specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

@@

@@

Now in order for the programmer to ensure that the {\em normalize} QRE is well
formed, she must define a canonizing function $f$ between two regular
languages and prove that $f$ is surjective and idempotent: the compiler cannot
verify this since this problem is undecidable in general. Consequently, we do
not use the $\normalize{R}{R'}{f}$ QRE in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonizer{}$ function for each QRE. Again, the \textit{permutation} combinator
gives the most interesting definition,
\begin{align*}\canonizer(\perm{c_1, \ldots, c_n}{c})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonizer(c_1)(r_1)) \cdot (\canonizer(c)(s_1)) \cdot \ldots \cdot
(\canonizer(c_n)(r_n))
\end{align*}
\noindent which as we have explained is defined so that the strings that match
$c_i, c$ are placed in the canonical permutation which is the identity
permutation before $\canonizer(c_i)$ and $\canonizer(c)$ are applied
respectively.

@@

The denotation of a lens $\ell$ is $\llbracket \ell \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell
\rrbracket$, then $\ell$ maps between $s_1$ and $s_2$.
\begin{align*}
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket const(r, s) \rrbracket &= \{(r, s)\}\\
\llbracket swap(\ell, \ell') \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \cdot \ell' \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \sep \ell' \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell \rrbracket \text{ or } (s, t) \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}\\
\llbracket \ell_2 \circ \ell_1 \rrbracket &= \{(r, t) \sep \text{there exists }s
\text{ such that } (r, s) \in \llbracket \ell_1 \rrbracket \text{ and } (s, t)
\in \llbracket \ell_2 \rrbracket\}
\end{align*}
The $\mathit{const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the backward
direction. $\mathit{id}(R)$ applies the identity function to the source and target
in $\mathcal{L}(R)$ in both directions. The composition lens $\ell' \circ \ell$
applies $\ell$ followed by then $\ell'$ to the source in the forward direction,
and applies $\ell'$ followed by $\ell$ to the target in the backward direction.
The lens $\ell \cdot \ell'$ first splits the string $s$ into $s_1$ and $s_2$,
applies $\ell$ and $\ell'$ to $s_1$ and $s_2$ to get $t_1$ and $t_2$
respectively, then concatenates $t_1$ and $t_2$ and returns $t_1 \cdot t_2$ as
the final result in the forward direction. $\ell \cdot \ell'$ operates
similarly in the backward direction, but with $s, s_1$ and $s_2$ substituted
for $t, t_1$ and $t_2$. The $\mathit{swap} \; (\ell, \ell')$ lens operates
like $\ell \cdot \ell'$, except that it swaps $t_1$ and $t_2$ before
concatenating the two for a final result of $t_2 \cdot t_1$ in the forward
direction. In the backward direction, $\mathit{swap}(\ell, \ell')$ first undoes
the swap, then proceeds as expected. The $\ell \; | \; \ell'$ lens
chooses to apply $\ell$ or $\ell'$ depending on whether the source
(resp. target) data is matched by $\ell$ or $\ell'$ in the forward (resp.
backward direction). The $\ell^*$ lens splits the string $s$ into strings $s_1,
\ldots, s_n$, applies $\ell$ to each $s_i$ to get $t_i$, and then concatenates
each of the $t_i$'s for a final result of $t_1 \cdot \ldots \cdot t_n$ in the
forward direction. $\ell^*$ operates similarly in the backward direction, but
with $s, s_i$ substituted for $t, t_i$.

@@

\begin{lstlisting}
let bib_to_end_q : (lens in bibtex $\Leftrightarrow$ endnote)  = rquot (lquot(bibtex, bib_to_end), endnote)
\end{lstlisting}

@@

In general, checking that $\canonizer(c') = \canonizer(c'')$ is an undecidable
task, especially because of the $\normalize{R}{R'}{f}$ QRE. Therefore, our
actual implementation of the composition rule for QRE lenses follows Boomerang's
approach, which is to define the composition on QRE lenses if and only if
$c'$ and $c''$ are syntactially equal to $id(R)$ and $id(R')$ for some $R$ and
$R'$ satisfying $\mathcal{L}(R) = \mathcal{L}(R')$.

@@

\begin{theorem}\label{normal form}
If there is a derivation $q : c \Leftrightarrow c'$, then there exists a
bijective lens $\ell : K(c) \Leftrightarrow K(c')$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c')
\end{align*}
\end{theorem}
\begin{proof}
We elide a full proof of this statement, since the proof is attained via a
straightforward induction on the derivation $q : c \Leftrightarrow c$ (the
full details can be found in the appendix {\sam{which I need to write}}). Since
bijective lenses are closed under strongly unambiguous regular operators, the
tricky case is composition. The composition case is more involved since we need
to show that we can get rid of the canonizers in the middle:

@@

\fi
\section{Synthesizing QRE Lenses}
\label{synth}
Our goal in this section is to demonstrate how to synthesize a quotient lens
$q: c \Leftrightarrow c'$ from the QREs $c, c'$ and a set of input-output pairs
$\{(x_1, y_1), \ldots, (x_n, y_n)\}$ where the $x_i$'s are in $W(c)$ and the
$y_i$'s are in $W(c')$. We wish $q$ to map the equivalence class of $x_i$ to the
equivalence class of $y_i$ and vice versa:
\begin{align*}
q.get(x_i) &\equiv_{c'} y_i \text{, and}\\
q.put(y_i) &\equiv_c x_i
\end{align*}
